---
title: "Translation Syntax Example"
---

```{r setup, include=FALSE}
# set the header {r} for all code chunks
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

##### Package Deposit
```{r warning = F}
## install.packages("descr")
library(descr) # freq()
## install.packages("ggplot2")
library(ggplot2)
```

### **Data Management**

##### create a data set with categorical and numeric variables and save as .csv
```{r}
# numerical variable
v1 <- runif(100, 0, 7.5) # generate 10 random numbers between 0 and 7.5
v2 <- runif(100, 7.5, 15)

# categorical variable
v3 <- sample(c("x", "y", "z", "n"), size = 100, replace = T) # randomly sample 10 observations out of x, y, z
v4 <- sample(c("a", "b", "c", "n"), size = 100, replace = T)
v5 <- sample(c("p", "q", "n"), size = 100, replace = T)

# combine variables into a dataframe
ie.df <- data.frame(v1, v2, v3, v4, v5) 

# save as csv
write.table(ie.df, file = "C:\\Users\\tonyr\\Downloads\\QAC201Z_Material\\example.csv", sep = ",", row.names = F)
```

##### sort the data
```{r}
# sort by ascending v1
ie.df <- ie.df[order(ie.df$v1, decreasing = F), ]
```

##### selecting observations
```{r}
# subset observations with x
ie.x <- ie.df[ie.df$v3 == "x", ]

# subset observations with x and a
ie.xa <- ie.df[ie.df$v3 == "x" & ie.df$v4 == "a", ]
```

##### Deal with missing data
```{r}
# label missing data with NA
## v3 and v4 "n" codes for NA
ie.df$v3[ie.df$v3 == "n"] <- NA
ie.df$v4[ie.df$v4 == "n"] <- NA
ie.df$v5[ie.df$v5 == "n"] <- NA

# subset observations without NA values in any variables
na.omit(ie.df)

# subset observations without NA values in selected variable
ie.df[!is.na(ie.df$v3),]
```

##### collapse response into a new variable
```{r}
# bin v1 into 4 sections 
ie.df$v6[ie.df$v1 < 2] <- 1
ie.df$v6[ie.df$v1 >=2 & ie.df$v1 < 4] <- 2
ie.df$v6[ie.df$v1 >= 4 & ie.df$v1 < 6] <- 3
ie.df$v6[ie.df$v1 >= 6 & ie.df$v1 < 8] <- 4

# combine v3 and v4 into v7
ie.df$v7[is.na(ie.df$v3) | is.na(ie.df$v4)] <- NA
ie.df$v7[ie.df$v3 == "x" & ie.df$v4 == "a"] <- "xa"
ie.df$v7[ie.df$v3 == "x" & ie.df$v4 == "b"] <- "xb"
ie.df$v7[ie.df$v3 == "x" & ie.df$v4 == "c"] <- "xc"
ie.df$v7[ie.df$v3 == "y" & ie.df$v4 == "a"] <- "ya"
ie.df$v7[ie.df$v3 == "y" & ie.df$v4 == "b"] <- "yb"
ie.df$v7[ie.df$v3 == "y" & ie.df$v4 == "c"] <- "yc"
ie.df$v7[ie.df$v3 == "z" & ie.df$v4 == "a"] <- "za"
ie.df$v7[ie.df$v3 == "z" & ie.df$v4 == "b"] <- "zb"
ie.df$v7[ie.df$v3 == "z" & ie.df$v4 == "c"] <- "zc"

# sum up v1 and v2 into v7
ie.df$v8 <- ie.df$v1 + ie.df$v2
```

##### rename variable
```{r}
names(ie.df)[names(ie.df) == "v8"] <- "v1_v2"
names(ie.df)[names(ie.df) == "v7"] <- "v3_v4"
names(ie.df)[names(ie.df) == "v6"] <- "v1_level"
```

##### label variable values
```{r}
levels(ie.df$v1_level) <- c("lowest", "low", "high", "highest")
```


### **Univariate Analysis**

##### **Categorical Variables**
```{r}
# frequency
#library(descr)
freq(as.ordered(ie.df$v3_v4))

# frequency barplot
#library(ggplot2)
ggplot(data = na.omit(ie.df)) + # na.omit() removes any observations containing NA
  geom_bar(aes(x = v3_v4)) +
  ggtitle("v3_v4 frequency barplot")
```

##### **Numeric Variables**
```{r}
# mean and standard deviation
mean(ie.df$v1_v2, na.rm = T)
sd(ie.df$v1_v2, na.rm = T)

# histogram
ggplot(data = ie.df)+
  geom_histogram(aes(x = v1_v2))+
  ggtitle("v1_v2 histogram")
```


### **Bivariate Analysis**

##### **Categorical-Categorical** (crosstabs)
```{r}
# crosstabs
table(ie.df$v3, ie.df$v4)
prop.table(table(ie.df$v3, ie.df$v4), 2) # col proportions
prop.table(table(ie.df$v3, ie.df$v4), 1) # row proportions
prop.table(table(ie.df$v3, ie.df$v4)) # cell proportions

# for binary categorical response
graph_data <- data.frame(v3 = na.omit(ie.df)$v3, 
                         v5 = ifelse(na.omit(ie.df)$v5 == "p", 0, 1))
ggplot(data = graph_data)+
  stat_summary(aes(x=v3, y = v5), fun = "mean", geom = "bar")+
  ylab("poriton of responses that are '1 (aka. q)'")+
  ggtitle("percentage of q within v3 groups")

# stacked barplot for mutli-level categorical responses
graph_data <- data.frame(v4 = c(rep("a", 3), rep("b", 3), rep("c", 3)),
                         v3 = rep(c("x", "y", "z"), 3),
                         value = as.numeric(table(ie.df$v3, ie.df$v4)))

ggplot(data = graph_data, aes(fill = v4, x = v3, y = value))+
  geom_bar(position = "stack", stat = "identity")+
  ylab("number of observations")+
  ggtitle("number of subjects at each response level within each group")
```


##### **Numeric-Categorical** (means by group)
```{r}
# mean by group
by(ie.df$v1, ie.df$v3_v4, mean, na.rm = T)
# standard deviation by group
by(ie.df$v1, ie.df$v3_v4, sd, na.rm = T)
# number of observations in group
by(ie.df$v1, ie.df$v3_v4, length)

# barplot
ggplot(data = na.omit(ie.df))+
  stat_summary(aes(x = v3_v4, y = v1), fun = "mean", geom = "bar")+
  ggtitle("mean barplot")

# boxplot
ggplot(data = na.omit(ie.df))+
  geom_boxplot(aes(x = v3_v4, y = v1))+
  ggtitle("boxplot")
```

##### **Numeric-Numeric**
```{r}
ggplot(data = na.omit(ie.df), aes(x = v1, y = v1_v2))+
  geom_point()+
  geom_smooth(method = "lm")+
  ggtitle("Correlation between v1_v2 and v1")
```


### **Multivariate Analysis** (Bivariate by subpopulation)

##### **Categorical-Categorical**
```{r}
# for binary categorical response
graph_data <- data.frame(v3 = na.omit(ie.df)$v3, 
                         v5 = ifelse(na.omit(ie.df)$v5 == "p", 0, 1),
                         v4 = na.omit(ie.df)$v4)
ggplot(data = graph_data)+
  stat_summary(aes(x=v3, y = v5), fun = "mean", geom = "bar")+
  facet_grid(.~v4)+
  ylab("poriton of responses that are '1 (aka. q)'")+
  ggtitle("percentage of q within v3 groups")

# stacked barplot for mutli-level categorical responses
graph_data <- data.frame(v4 = rep(c(rep("a", 3), rep("b", 3), rep("c", 3)), 2),
                         v3 = rep(c("x", "y", "z"), 6),
                         value = c(as.numeric(table(ie.df[ie.df$v5 == "p",]$v3, ie.df[ie.df$v5 == "p",]$v4)), 
                                   as.numeric(table(ie.df[ie.df$v5 == "q",]$v3, ie.df[ie.df$v5 == "q",]$v4))),
                         v5 = c(rep("p", 9), rep("q", 9)))

ggplot(data = graph_data)+
  geom_bar(position = "stack", aes(fill = v4, x = v3, y = value), stat = "identity")+
  facet_grid(.~v5)+
  ylab("number of observations")+
  ggtitle("number of subjects at each response level within each group")
```

##### **Numeric-Categorical**
```{r}
# mean by group 
ftable(by(ie.df$v1_v2, list(ie.df$v3_v4, ie.df$v5), mean, na.rm = T))

# boxplot
ggplot(data = na.omit(ie.df)) +
  geom_boxplot(aes(x = v3_v4, y = v1_v2))+
  facet_grid(.~v5)+
  ggtitle("")
```

##### **Numeric-Numeric**
```{r}
ggplot(data = na.omit(ie.df), aes(x = v1, y = v1_v2))+
  geom_point()+
  geom_smooth(method = "lm")+
  facet_grid(.~v5)
```


### **Hypothesis Testing**

#### **Categorical-Categorical ($\chi^2$)**
##### Bivariate
```{r}
# chisquare
chisq.test(na.omit(ie.df)$v4, na.omit(ie.df)$v3)

# chisquare cell counts
chisq.test(na.omit(ie.df)$v4, na.omit(ie.df)$v3)$observed

# posthoc (for more than two levels)
source("https://raw.githubusercontent.com/PassionDrivenStatistics/R/master/ChiSquarePostHoc.R")
chisq.post.hoc(chisq.test(na.omit(ie.df)$v4, na.omit(ie.df)$v3)$observed,
               popsInRows = F,
               control = "bonferroni")

# Pearson Residuals
chisq.test(na.omit(ie.df)$v4, na.omit(ie.df)$v3)$residuals
```
##### Multivariate
```{r}
by(na.omit(ie.df), na.omit(ie.df)$v5, function(x) list(chisq.test(x$v4, x$v3),
                                                       chisq.test(x$v4, x$v3)$observed,
                                                       prop.table(chisq.test(x$v4, x$v3)$observed, 2)))
```

#### **Numeric-Categorical (ANOVA)**
##### Bivariate
```{r}
# anova
summary(aov(v1_v2 ~ v3_v4, data = na.omit(ie.df)))

# posthoc
TukeyHSD(aov(v1_v2 ~ v3_v4, data = na.omit(ie.df)))
```
##### Multivariate
```{r}
by(na.omit(ie.df), na.omit(ie.df)$v5, function(x) list(aov(v1_v2 ~ v3_v4, data = x),
                                                       summary(aov(v1_v2 ~ v3_v4, data = x)),
                                                       TukeyHSD(aov(v1_v2 ~ v3_v4, data = x))))
```


#### **Numeric-Numeric (Pearson Correlation)**
##### Bivariate
```{r}
cor.test(na.omit(ie.df)$v1_v2, na.omit(ie.df)$v1)
```
##### Multivariate
```{r}
by(na.omit(ie.df), na.omit(ie.df)$v5, function(x) cor.test(x$v1_v2, x$v1))
```


### **Regression**

##### **linear regression (numeric response)**
```{r}
# simple
summary(lm(v1_v2 ~ v1, data = na.omit(ie.df)))

# multiple
summary(lm(v1_v2 ~ v1 + factor(v5), data = na.omit(ie.df)))

# multiple with interaction terms
summary(lm(v1_v2 ~ v1 + factor(v5) + v1*factor(v5), data = na.omit(ie.df)))
```

##### **logistic regression (categorical response)**
```{r}
# simple
summary(glm(factor(v5) ~ v1, data = na.omit(ie.df), family = "binomial")) # p-value
exp(glm(factor(v5) ~ v1, data = na.omit(ie.df), family = "binomial")$coefficients) # odd-ratios
exp(confint(glm(factor(v5) ~ v1, data = na.omit(ie.df), family = "binomial"))) # confidence interval of odd ratios


# multiple
summary(glm(factor(v5) ~ v1 + factor(v3), data = na.omit(ie.df), family = "binomial")) # p-value
exp(glm(factor(v5) ~ v1 + factor(v3), data = na.omit(ie.df), family = "binomial")$coefficients) # odd-ratios
exp(confint(glm(factor(v5) ~ v1 + factor(v3), data = na.omit(ie.df), family = "binomial"))) # confidence interval of odd ratios

# multiple with interaction terms
summary(glm(factor(v5) ~ v1 + factor(v3) + v1*factor(v3), data = na.omit(ie.df), family = "binomial")) # p-value
exp(glm(factor(v5) ~ v1 + factor(v3) + v1*factor(v3), data = na.omit(ie.df), family = "binomial")$coefficients) # odd-ratios
exp(confint(glm(factor(v5) ~ v1 + factor(v3) + v1*factor(v3), data = na.omit(ie.df), family = "binomial"))) # confidence interval of odd ratios
```

```{r}

```